When Proximity Isn't Enough: Network Segregation and SES Bias in Referrals

### 2025-08-07 writing/manuscript.tex
v1.2

\section*{Abstract}
The share of high-SES connections in one's network is a strong correlate of labor market income. We investigate whether SES biases in referral selection exacerbate differences high-SES connection shares. We conduct a lab-in-the-field experiment with 734 Colombian university students who make incentivized referrals from their enrollment networks. Randomizing participants between performance-only incentives and performance plus a fixed bonus for referral recipients, we find that referrals go to high-performing peers with whom they take many courses together, regardless of incentives. While low-SES referrers exhibit strong in-group preferences, middle- and high-SES referrers show no bias toward other groups and refer along the network shares of each SES groups. We find that network segregation, driven by program selection based on SES, limits cross-SES referral opportunities for even without an explicit SES bias. These suggest institutional policies promoting cross-SES contact are key for reducing SES-based inequalities.
\medskip \\
\textbf{JEL Classification:} C93, J71, D85, Z13 \\
\textbf{Keywords:} inequality, economic mobility, peer networks, class discrimination, homophily

@@

We find that referrals consistently go to higher-performing peers with high connection intensity (14 vs. 4 courses), regardless of the conditions and the exam area. Pooling across these, we find that SES bias in referral selection is primarily driven by low-SES participants exhibiting in-group preferences: Controlling for network composition, low-SES referrers are 45\% more likely to refer other low-SES peers  and 44\% less likely to refer high-SES relative to middle-SES peers. In contrast, middle- and high-SES referrers show no bias toward other groups.

@@

A subset of the literature focuses on SES-based differences in social capital and network formation \citep{chetty2022na,engzell2025ajos,michelman2022tqjoe}, with connection intensity \citep{gee2017jole,kramarz2014res,wang2013aejae,sterling2014ms} and homophily \citep{mcpherson2001birds,currarini2009e,jackson2022,bolte2024jle,montgomery_social_1991} driving differences across groups. Based on the pioneering work of \citet{currarini2010pnas}, we contribute by identifying two different types of homophily, and separate whether differential outcomes stem from network composition (who you know) versus taste-based biases (who you choose to interact with). Our findings suggest that rather than focusing on taste-based biases,  implementing mixed-program courses to increase across-SES connection intensity should be a clear policy goal in order to reduce SES-based network segregation (see e.g., for an institutional intervention by \citet{rohrer2021po}).

Methodologically, we contribute to the literature on job referral experiments. This literature provides causal evidence on why referrals in the labor market are prevalent,\footnote{Referrals solve frictions in the search and matching process and benefit both job-seekers and employers \citep{topa2019social}. Referral candidates tend to get hired more often, have lower turnover, and earn higher wages \citep{obukhova2013ms,brown2016informal,dustmann2016referral}.} finding that performance-based incentives bring in qualified candidates otherwise not identified by demographic characteristics \citep{beaman_who_2012,pallais_why_2016,friebel2023employee,witte2021workers}, and the consequences of relying upon referral hiring, which come at the cost of disadvantaging certain groups \citep{beaman_job_2018,hederos2025gender}. We extend this literature by causally evaluate the effect of a sizeable monetary bonus for the referral candidate and exploring SES biases in referral selection.

@@

What are the diversity-related consequences of SES-driven differences across networks? In terms of network compositions, participants could connect with other SES groups at different rates than would occur randomly depending on their own SES. Figure \ref{connections_to_low_ses} and Figure \ref{connections_to_high_ses} illustrate the average network shares conditional on referrer SES respectively for low- and high-SES.\footnote{For sake of brevity we omit middle-SES from this exposition. For the complete relationship, see Appendix Figure \ref{connections_by_ses_app}.} We observe modest deviations from university-wide SES shares in enrollment networks: Low-SES referrers have on average 38.4\% low-SES peers compared to the university average of 34.3\%, while high-SES referrers have 20.4\% high-SES connections compared to the university average of 15.3\%.

@@

\begin{figure}[H]
  \centering
  \caption{Network shares of SES groups}\label{availability_lses}
  \begin{subfigure}[t]{0.4\textwidth} 
    \centering
      \includegraphics[width=\linewidth]{../figures/connections_to_low_ses.png}
      \caption{Low-SES network share}  
      \label{connections_to_low_ses}
  \end{subfigure}
  \begin{subfigure}[t]{0.4\textwidth} 
    \centering
      \includegraphics[width=\linewidth]{../figures/connections_to_high_ses.png}
      \caption{High-SES network share}
      \label{connections_to_high_ses}
  \end{subfigure}
  \caption*{\footnotesize\textit{Note:} Figures illustrate the average network shares of low- and high- SES peers conditional on referrer SES. Horizontal lines plot the university-wide shares of SES groups (Low: 35\%, High: 14\%). While the share of low-SES peers in the network decreases as referrer SES increases, the share of high-SES peers in the network increases. Error bars represent 95\% confidence intervals. }
\end{figure}

While same-SES students are connected more often with each other, so far we only consider the average the number of courses taken together with network members. What are the diversity implications of increased connection intensity between students? As students take more courses together with peers, the share of same-SES peers in the networks of low- and high-SES increases while the share of middle-SES declines (see Figure \ref{share_same_ses_by_tie}). Both increases are substantial, amounting to 50\% for high-, and 30\% for low-SES beyond 15 courses together. Considering that beyond 5 courses taken together network members are almost entirely within the same program, program selection may have strong consequences for explaining the network segregation in our setting.

@@

Put differently, in an environment where 1 out of 3 students are low-SES, the chance that low-SES are considered for a referral by high-SES at random is already less than 1/6. This stark disparity reveals that low-SES and high-SES students operate in practically separate social worlds within the same university, despite formal contact opportunities to meet as equal-status students. While referral selection being driven by availability and performance is positive, network segregation has such a large impact on diversity. We now explore program selection as a key driver.

\begin{figure}[H]
  \centering
  \caption{Network shares of SES groups above median connection intensity}
  \begin{subfigure}[t]{0.4\textwidth} 
    \centering
      \includegraphics[width=\linewidth]{../figures/connections_to_low_ses_tie12.png}
      \caption{Low-SES network share}  
      \label{connections_to_low_ses_tie12}
  \end{subfigure}
  \begin{subfigure}[t]{0.4\textwidth} 
    \centering
      \includegraphics[width=\linewidth]{../figures/connections_to_high_ses_tie12.png}
      \caption{High-SES network share}
      \label{connections_to_high_ses_tie12}
  \end{subfigure}
  \caption*{\footnotesize\textit{Note:} Figures illustrate the average network shares of low- and high- SES peers conditional on referrer SES above the median number of courses taken together. Horizontal lines plot the university-wide shares of SES groups (Low: 35\%, High: 14\%). While the share of low-SES peers in the network decreases as referrer SES increases, the share of high-SES peers in the network increases. Error bars represent 95\% confidence intervals. }
\end{figure}

\subsection{Program selection and SES diversity}

Academic programs at this university use cost-based pricing, and typically less than 5\% of students receive any kind of scholarship. Based on this, we first calculate how much every undergraduate program at the university is expected to cost students per year (see Figure \ref{fees_with_boxplot}). Considering that net minimum monthly wage stands at \$200 and the average Colombian salary around \$350, the cost differences between programs are large enough to make an impact on program selection. Is it the case that SES groups select into programs with financial considerations?

@@

We then examine how SES groups are distributed across programs to identify evidence of SES-based selection (see Figure \ref{ses_distribution_by_fees}). Indeed, low-SES students select into more affordable programs, followed by middle-SES students. High-SES students sort almost exclusively into above-average costing programs, with a third selecting into medicine and creating a very skewed distribution. The distributions are significantly different across all pairwise comparisons: low-SES vs. middle-SES (Kolmogorov-Smirnov test $D=33.89$, $p<0.001$), low-SES vs. high-SES ($D=31.31$, $p<0.001$), and middle-SES vs. high-SES ($D=31.31$, $p<0.001$). With this finding, program selection could be the reason why low- and high-SES networks tend to segregate as the number of courses taken increases. The next section characterizes the referrals, and we will return to the diversity implications of program selection once we propose an understanding of how referrals were made.

@@

Does the number of courses taken together have an independent effect that goes beyond identifying peers in the same academic program? To evaluate this question we leverage our administrative data, and identify peers within the same program: In each individual network we observe the participant-specific academic program for the participant making the referral and alternative-specific academic program for each referral candidate. We add this new variable in our specification and describe our findings in Table \ref{tab:ses-heterogeneity-program}. Being in the same academic program has a substantial positive effect on referral likelihood, with coefficients ranging from 1.257 to 2.198 across all referrer SES groups. This confirms that program affiliation serves as a strong predictor of referral decisions, reflecting increased familiarity. Our comparison of interest is the point estimate for the standardized number of courses taken. Across all three referrer groups, the standardized number of courses taken together maintains its statistical significance after controlling for same program membership. The coefficient magnitudes are expectedly smaller compared to specifications without program controls (ranging from 0.688 to 0.930) as the newly added variable is a moderator: Matching academic programs leads to taking more courses together. The remaining estimates in our model remain robust to the inclusion of the same-program variable with little change in point estimates. The persistence of statistical significance (all $p < 0.001$) suggests that the number of courses taken together has an independent effect on referral decisions. To sum, our measure of connection intensity seems to capture meaningful social interaction patterns that lead to referrals, and go beyond simply identifying matching academic programs.

@@

\section{Conclusion} \label{Conclusion}
We investigate whether SES biases in referral selection stem from taste-based preferences in choosing an SES group over others or network segregation. Through a lab-in-the-field experiment with 734 university students making incentivized referrals from complete enrollment networks, we find that institutional factors dominate individual preferences.

Our key findings are threefold. First, referral patterns remain unchanged across different incentive structures: participants consistently select high-performing peers with a high number of courses taken together regardless of whether referral recipients receive additional compensation. Second, we find an SES bias is that is asymmetric and limited. While low-SES referrers exhibit strong in-group preferences,  middle- and high-SES referrers show no bias toward other groups. Third, network segregation driven by cost-based program selection explains most referral patterns. At typical referral range measured by the number of courses taken together, low-SES and high-SES students have dramatically different choice sets, with high-SES networks containing only 15.7\% low-SES peers compared to 35\% university-wide.

These results have important policy implications. While universities expose low-SES students to higher-than-population shares of high-SES peers, segregation within institutions limits meaningful cross-SES interaction. Our findings suggest that institutional interventions promoting mixed-SES classrooms, rather than addressing individual biases, represent the most promising approach to reducing SES-based inequality in opportunity transmission. Future research should explore the causal effect of specific institutional interventions that aim to increase cross-SES interactions.

>>>>

\section*{Abstract}
The share of high-SES connections in one's network is a strong correlate of labor market income. We investigate whether SES biases in referral selection exacerbate differences high-SES connection shares. We conduct a lab-in-the-field experiment with 734 Colombian university students who make incentivized referrals from their enrollment networks. Randomizing participants between performance-only incentives and performance plus a fixed bonus for referral recipients, we find that referrals go to high-performing peers with whom they take many courses together, regardless of incentives. While low-SES referrers exhibit strong in-group preferences, middle- and high-SES referrers show no biases towards their own and other groups, referring along their network shares. We find that network segregation, driven by program selection based on SES, limits cross-SES referral opportunities for even without an explicit SES bias. These suggest institutional policies promoting cross-SES contact are key for reducing SES-based inequalities.
\medskip \\
\textbf{JEL Classification:} C93, J71, D85, Z13 \\
\textbf{Keywords:} inequality, economic mobility, peer networks, class discrimination, homophily

@@

We find that referrals consistently go to higher-performing peers with high connection intensity (14 vs. 4 courses), regardless of the conditions and the exam area. Pooling across these, we find that SES bias in referral selection is primarily driven by low-SES participants exhibiting in-group preferences: Controlling for network composition, low-SES referrers are 45\% more likely to refer other low-SES peers  and 44\% less likely to refer high-SES relative to middle-SES peers. In contrast, middle- and high-SES referrers show no biases toward their own or other groups.

@@

A subset of the literature focuses on SES-based differences in social capital and network formation \citep{chetty2022na,engzell2025ajos,michelman2022tqjoe}, with connection intensity \citep{gee2017jole,kramarz2014res,wang2013aejae,sterling2014ms} and homophily \citep{mcpherson2001birds,currarini2009e,jackson2022,bolte2024jle,montgomery_social_1991} driving differences across groups. Based on the pioneering work of \citet{currarini2010pnas}, we contribute by identifying two different types of homophily, and separate whether differential referral outcomes stem from network composition (who you know) versus taste-based biases (who you choose to interact with). Our findings suggest that structural factors impacting network composition, rather than taste-based SES biases, drive the differences in referral outcomes. Under this light, implementing mixed-program courses to increase across-SES connection intensity should be a clear policy goal in order to reduce SES-based network segregation.

Methodologically, we contribute to the literature on job referral experiments. This literature provides causal evidence on why referrals in the labor market are prevalent,\footnote{Referrals solve frictions in the search and matching process and benefit both job-seekers and employers \citep{topa2019social}. Referral candidates tend to get hired more often, have lower turnover, and earn higher wages \citep{obukhova2013ms,brown2016informal,dustmann2016referral}.} finding that performance-based incentives bring in qualified candidates otherwise not identified by demographic characteristics \citep{beaman_who_2012,pallais_why_2016,friebel2023employee,witte2021workers}, and the consequences of relying upon referral hiring, which come at the cost of disadvantaging certain groups \citep{beaman_job_2018,hederos2025gender}. We extend this literature by causally evaluating the effect of a sizeable monetary bonus for the referral candidate and exploring SES biases in referral selection.

@@

What are the diversity consequences of SES-driven differences across networks? In terms of network compositions, participants could connect with other SES groups at different rates than would occur randomly depending on their own SES. Figure \ref{connections_to_low_ses} and Figure \ref{connections_to_high_ses} illustrate the average network shares conditional on referrer SES respectively for low- and high-SES.\footnote{For sake of brevity we omit middle-SES from this exposition. For the complete relationship, see Appendix Figure \ref{connections_by_ses_app}.} We observe modest deviations from university-wide SES shares in enrollment networks: Low-SES referrers have on average 38.4\% low-SES peers compared to the university average of 34.3\%, while high-SES referrers have 20.4\% high-SES connections compared to the university average of 15.3\%.

@@

\begin{figure}[H]
  \centering
  \caption{Network shares of SES groups}\label{availability_lses}
  \begin{subfigure}[t]{0.4\textwidth} 
    \centering
      \includegraphics[width=\linewidth]{../figures/connections_to_low_ses.png}
      \caption{Low-SES network share}  
      \label{connections_to_low_ses}
  \end{subfigure}
  \begin{subfigure}[t]{0.4\textwidth} 
    \centering
      \includegraphics[width=\linewidth]{../figures/connections_to_high_ses.png}
      \caption{High-SES network share}
      \label{connections_to_high_ses}
  \end{subfigure}
  \caption*{\footnotesize\textit{Note:} Figures illustrate the average network shares of low- and high- SES peers conditional on referrer SES. Horizontal lines plot the university-wide shares of SES groups (Low: 34\%, High: 15\%). While the share of low-SES peers in the network decreases as referrer SES increases, the share of high-SES peers in the network increases. Error bars represent 95\% confidence intervals. }
\end{figure}

While same-SES students are connected more often with each other, so far we only consider the average the number of courses taken together with network members. What are the diversity implications of increased connection intensity between students? As students take more courses together with peers, the share of same-SES peers in the networks of low- and high-SES increases while the share of middle-SES declines (see Figure \ref{share_same_ses_by_tie}). Both increases are substantial, amounting to 50\% for high-, and 30\% for low-SES beyond 15 courses together.  
While it is known that students who take courses together have similar characteristics \citep{kossinets_origins_2009}, it is important to understand how increasing similarities in SES reflects on referral choice sets.

@@

Put differently, in an environment where 1 out of 3 students are low-SES, the chance that a low-SES peer is considered for a referral by high-SES is already less than 1/6. This stark disparity reveals that low-SES and high-SES students practically have separate networks within the same university, despite  the opportunities to meet as equal-status students. The network segregation makes cross-SES referrals structurally unlikely even without any taste-based SES biases. We now explore program selection that emerges as a key driver of this segregation.

\begin{figure}[H]
  \centering
  \caption{Network shares of SES groups above median connection intensity}
  \begin{subfigure}[t]{0.4\textwidth} 
    \centering
      \includegraphics[width=\linewidth]{../figures/connections_to_low_ses_tie12.png}
      \caption{Low-SES network share}  
      \label{connections_to_low_ses_tie12}
  \end{subfigure}
  \begin{subfigure}[t]{0.4\textwidth} 
    \centering
      \includegraphics[width=\linewidth]{../figures/connections_to_high_ses_tie12.png}
      \caption{High-SES network share}
      \label{connections_to_high_ses_tie12}
  \end{subfigure}
  \caption*{\footnotesize\textit{Note:} Figures illustrate the average network shares of low- and high- SES peers conditional on referrer SES above the median number of courses taken together. Horizontal lines plot the university-wide shares of SES groups (Low: 34\%, High: 15\%). While the share of low-SES peers in the network decreases as referrer SES increases, the share of high-SES peers in the network increases. Error bars represent 95\% confidence intervals. }
\end{figure}

\subsection{Program selection as a mechanism}

Academic programs at this university have cost-based pricing, and typically less than 5\% of students receive any kind of scholarship. Based on this, we first calculate how much every undergraduate program at the university is expected to cost students per year (see Figure \ref{fees_with_boxplot}). Considering that net minimum monthly wage stands at \$200 and the average Colombian salary around \$350, the cost differences between programs are large enough to make an impact on program selection. Is it the case that SES groups select into programs with financial considerations?

@@

To answer, we examine how SES groups are distributed across programs to identify evidence of SES-based selection (see Figure \ref{ses_distribution_by_fees}). Indeed, low-SES students select into more affordable programs, followed by middle-SES students. High-SES students sort almost exclusively into above-average costing programs, with a third selecting into medicine and creating a very skewed distribution. The distributions are significantly different across all pairwise comparisons: low-SES vs. middle-SES (Kolmogorov-Smirnov test $D=33.89$, $p<0.001$), low-SES vs. high-SES ($D=31.31$, $p<0.001$), and middle-SES vs. high-SES ($D=31.31$, $p<0.001$). These findings support the idea that program selection could be the reason why low- and high-SES networks tend to segregate as the number of courses taken increases. Financial constraints channel students into different academic programs, which in turn determine their classroom interactions and university social networks.

@@

Does the number of courses taken together have an independent effect that goes beyond identifying peers in the same academic program? To evaluate this question we leverage our administrative data, and identify peers within the same program: In each individual network we observe the participant-specific academic program for the participant making the referral and alternative-specific academic program for each referral candidate. We add this new variable in our specification and describe our findings in Table \ref{tab:ses-heterogeneity-program}. Being in the same academic program has a substantial positive effect on referral likelihood, with coefficients ranging from 1.257 to 2.198 across all referrer SES groups. This confirms that program affiliation serves as a strong predictor of referral decisions. Our comparison of interest is the point estimate for the standardized number of courses taken. Across all three referrer groups, the standardized number of courses taken together maintains its statistical significance after controlling for same program membership. The coefficient magnitudes are expectedly smaller compared to specifications without program controls (ranging from 0.688 to 0.930) as the newly added variable is a moderator: Matching academic programs leads to taking more courses together. The remaining estimates in our model remain robust to the inclusion of the same-program variable with little change in point estimates. The persistence of statistical significance (all $p < 0.001$) suggests that the number of courses taken together has an independent effect on referral decisions. To sum, our measure of connection intensity seems to capture meaningful social interaction patterns that lead to referrals, and go beyond simply identifying matching academic programs.

@@

\section{Conclusion} \label{Conclusion}
We investigate whether SES biases in referral selection stem from taste-based preferences in choosing an SES group over others or network segregation. Through a lab-in-the-field experiment with 734 university students making incentivized referrals from their unique enrollment networks, we find that institutional factors dominate individual preferences.

Our key findings are threefold. First, referral patterns remain unchanged across different incentive structures: participants consistently select high-performing peers with a high number of courses taken together regardless of whether referral recipients receive additional compensation. Second, we find an SES bias is that is asymmetric and limited. While low-SES referrers exhibit strong in-group preferences,  middle- and high-SES referrers show no bias toward their own and other groups. Third, network segregation driven by cost-based program selection explains most referral patterns. At typical referral range measured by the number of courses taken together, low-SES and high-SES students have dramatically different choice sets, with high-SES networks containing only 15.7\% low-SES peers compared to 34\% university-wide.

These results have important policy implications. While universities expose low-SES students to higher-than-population shares of high-SES peers, segregation within institutions limits meaningful interaction across SES. Our findings suggest that institutional interventions promoting cross-SES contact, represents a promising approach to reduce SES-based inequality in opportunity transmission. Future research should explore the causal effects of specific institutional interventions such as mixed seating \citep{rohrer2021po}, or cross-SES mentoring programs \citep{alan2025aer}, that increase interactions between with SES groups.
